Enhancing HiC data resolution with convolutional neural networks

### 2018-04-23 HiCPlus/tex/review.tex
finished report

%% import packages that have custom options
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[pagebackref=true]{hyperref}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{float}
\usepackage[inline]{enumitem}

%% Import general packages
\usepackage{
  amsmath, amssymb, amsfonts, nicefrac,
  algorithmic, textcomp, listings, url,
  graphicx, subfig, microtype,
  booktabs, longtable
}

@@

In our research, we have Hi-C data of
4 cell lines. One of which is sequenced
from a normal cell line and the other
three sequenced from cells afflicted
with three different malignancies.
Our purpose is to
compare them in terms of spatial
structure and find whether there is 
any difference in their 3D conformation
or not.
All 4 data that we have are sequenced
with low depth, resulting in relatively
low resolution. Therefore, we used the
HiCPlus model in \cite{zhang2018enhancing}
to enhace the contrast of our data.
We also have access to a high-resolution
data with much higher resolution which
is sequenced from exactly the same cell
as the normal low-resolution data that
we have. Our purpose to train the model
by using the low- and high-resolution data
of normal cell and then apply it on the 
other three cells in order to improve
their contrast.

@@

    \frac{1}{784}\sum_{i=1}^{784}{\hat{y}_i - y_i}

@@

@@

    \includegraphics[width=\textwidth]{mit.png}
    \caption{}
        Malignant Cell 1: B-ALL
        \includegraphics[width=\textwidth]{all.png}
        \vspace{10pt}
        Malignant Cell 2: RL
        \includegraphics[width=\textwidth]{rl.png}
        \vspace{10pt}
        Malignant Cell 3: CALL4
        \includegraphics[width=\textwidth]{call4.png}
    \caption{}

@@

We trained the model using low-resolution 
 GM06990 and high-resolution
 GM12878 data, both of which are sequence from
 the same cell. We then used the other three
 data that we have (corresponding to three
 cancerous cells) as input to the model to 
 improve contrast.
The data that we uses is described in table
\ref{tab:data}
\section{Strengths and Weaknesses and Future Work}
This method can be considered as a noise reduction method
for Hi-C contact matrices. 
For example Balance Network Deconvolution,
proposed by \cite{feizi2013network}
assumes that the observed graph $G_{obs}$ is a summation of its
direct graph $G_{dir}$ and some indirect terms as follows:
\begin{equation}
        \label{indirect_effects}
            G_{obs} = G_{dir} + G_{dir}^2 + G_{dir} + ...
\end{equation}
Just as all currently available
normalization and noise reduction approaches this method
also relies on assumptions. The major assumption of this
approach is that CNNs can predict values in high-resolution
matrix from the surrounding neighborhood in low-resolution
matrix. The difference of this assumption from other 
approaches is that it is easier to put the assumption to
statistical tests. Authors of \cite{zhang2018enhancing}
came to the conclusion that the assumption holds true for
a down-sampled version of high-resolution data. Whether
this is the case for an independent low-resolution data
remains to be investigated in future works.
\section{Resources}
\textbf{Source code for the project:}\\
\url{https://github.com/rasoolianbehnam/watson/tree/master/HiCPlus}\\
\textbf{Link to the low-resolution Hi-C datasets:}\\
\url{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/contact_file/}\\
\textbf{Link to high-resolution Hi-C dataset:}\\
\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525}

>>>>

%% import packages that have custom options
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[pagebackref=true]{hyperref}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{float}
\usepackage{graphicx}
\usepackage[caption = false]{subfig}
%% Import general packages
\usepackage{
  amsmath, amssymb, amsfonts, nicefrac,
  algorithmic, textcomp, listings, url,
  graphicx, subfig, microtype,
  booktabs, longtable
}

@@

Our final purpose is to be able to
compare the three leukemic
cells with a normal cell
in terms of spatial
structure and find whether there is 
any difference in their 3D conformation
or not.
In our research, we have 5 Hi-C data sets,
belonging to
4 cell lines. Two data sets are  sequenced
from a normal cell line, one 
with high resolution (GM12878) and the
other with low resolution (GM06990)
The other three data sets are
sequenced from the same
cell lines with with three different
type of leukemia.

As mentioned above, four of the data 
sets we have are sequenced
with low depth, resulting in relatively
low resolution. 
We also have access to a high-resolution
data with much higher resolution which
is sequenced from exactly the same cell
as the normal low-resolution data that
we have. 
Therefore, we used the
HiCPlus model in \cite{zhang2018enhancing}
to enhace the contrast of our data.
We trained  the model
by using the low- and high-resolution data
of normal cells and then applied it to the 
other three cells in order to improve
their contrast.
A summary of the data and how
they are used can be 
found in table \ref{tab:data}.

@@

    \frac{1}{784}\sum_{i=1}^{784}({\hat{y}_i - y_i})^2

@@

We used simple gradient descent optimizer with learning
rate of \texttt{1e-5} to train the model. The choise of
optimizer did not make a great difference in the results.
Our results from using Adam optimizier also were the same
as SGD.

@@

\begin{figure}
        \includegraphics[width= 6in]{mit.png}
        \caption{Results of training the data on
        Normal cells. There is a total of almost
        6000 loci in a contact matrix. For the
        sake of demonstration, we only show the
        firt 500 loci. The left contact matrix is the 
        original low-resolution data. The middle one 
        is the high resolution data. and the rightmost
        picture is the result of training the model.}
        \label{fig:train}
\end{figure}
\begin{figure}
        \subfloat[]{\includegraphics[width= 4in]{all.png}}\\
        \subfloat[]{\includegraphics[width= 3.9in]{rl.png}}\\
        \subfloat[]{\includegraphics[width= 4in]{call4.png}}\\
        \caption{Results of running the model on
        cancer cells. For the
        sake of demonstration, we only show the
        firt 500 loci. The left contact matrices 
        are the 
        original low-resolution data and the rightmost
        pictures are the result of model predictions.}
        \label{fig:predict}
\end{figure}

@@

We trained the model using low-resolution 
 GM06990 and high-resolution
 GM12878 data, both of which are sequence from
 the same cell. We then used the other three
 data that we have (corresponding to three
 cancerous cells) as input to the model to 
 improve contrast.
The data that we uses is described in table
\ref{tab:data}

The model was ran for 30 epochs. 
The results of training can be found in figure
\ref{fig:train}. As can be seen the results
are smoother and demonstrate higher contrast
than the original model.
Figure \ref{fig:predict} show prediction results
from the model for the three cancer cells. The
results also demonstrat higher contrast compared
to original contact matrices.

\section{Strengths and Weaknesses and Future Work}
This method can be considered as a noise reduction method
for Hi-C contact matrices. 
Just as all currently available
normalization and noise reduction approaches this method
also relies on assumptions. 
For example Balance Network Deconvolution,
proposed by \cite{feizi2013network}
assumes that the observed graph $G_{obs}$ is a summation of its
direct graph $G_{dir}$ and some indirect terms 
$G_{obs} = G_{dir} + G_{dir}^2 + G_{dir} + ...$ .
The major assumption of this
approach is that CNNs can predict values in high-resolution
matrix from the surrounding neighborhood in low-resolution
matrix. The difference of this assumption from other 
approaches is that it is easier to put the assumption to
statistical tests. Authors of \cite{zhang2018enhancing}
came to the conclusion that the assumption holds true for
a down-sampled version of high-resolution data. Whether
this is the case for an independent low-resolution data
remains to be investigated in future works.
\section{Resources}
\textbf{Source code for the project:}\\
\url{https://github.com/rasoolianbehnam/watson/tree/master/HiCPlus}\\
\textbf{Link to the low-resolution Hi-C datasets:}\\
\url{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/contact_file/}\\
\textbf{Link to high-resolution Hi-C dataset:}\\
\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525}
